\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts}
\renewcommand\vec[1]{\mathbf{#1}}
\usepackage{listings}
\usepackage[table]{xcolor}
\usepackage{framed,comment}
\newcommand\vx {\mathbf{x}}
\newcommand\vy {\mathbf{y}}
\newcommand\vb{\mathbf{b}}
\newcommand\vc{\mathbf{c}}
\newcommand\red[1]{\textcolor{red}{#1}}
\specialcomment{solution}{\bigskip\begin{leftbar}\par\noindent\textbf{Solution.} }{\end{leftbar} }
\excludecomment{solution}
\begin{document}

\begin{tabular}{l}
	\textbf{CSCI 5654-Fall16}: Assignment \#4 (Reading: 6.1-6.3,
  7.1, 12).                                                  \\
	\textbf{Due Date:} Monday, Oct 24, 2016 (before class)            \\
	\textbf{In-class:} Assignment should be submitted on paper -- no emails. \\
	\textbf{Distance Students:} Assignment may be submitted on paper or
	by email.                                                                \\[10pt]

	\textbf{Your Name:} \phantom{Supercalifragilisticexpialidocius Smith}    \\
	\hline
	                                                                         \\[10pt]
\end{tabular}

\noindent\textbf{P1. (10 points)}  Express the problems below as linear
programs by adding suitable additional variables.

\bigskip

\noindent\textbf{(A)} $\min_{x_1, x_2, x_3 \in \mathbb{R}^3}\ \max( 2 x_1 + 3 x_2 - 5 x_3,\ x_1,\ x_2,\ 2
x_1 - x_2 + x_3)$.

\medskip 

\noindent\textbf{(B)} $\min_{x_1, x_2, x_3 \in \mathbb{R}^3}\ \left(
  |x_1 + x_2| + | x_2 - x_3 | + | x_3 - x_1 | + | x_1 + x_2 + x_3 |
\right) $.

\medskip

\noindent\textbf{(C)} \[ \begin{array}{rl}
\min\ & \max\left( | x_1 | ,\ |x_2| ,\ |x_3|,\ |x_1+
  x_2 | \right) \\
 \mathsf{subj. to}\ & x_1 - x_2 \leq 5 \\
&  x_2 \leq 3 \end{array} \]

\bigskip

\noindent\textbf{P2. (15 points)} The attached spreadsheet provides
values of insulin input $u(t)$ and the blood glucose levels $G(t)$ 
for a patient with type-1 diabetes. Our goal is to use an 
ARMAX model that predicts the output $G(t+1)$ as a function of the
past glucose and insulin values:

\[ G(t+1) = a_0 G(t) + a_1 G(t-1) + \cdots + a_{10} G(t - 10) + b_0
  u(t) + \cdots + b_{5} u(t-5) \]

Use linear regression using $L_1$ norm on the given data to 
learn the arma model
coefficients $a_0, a_1, \ldots, b_0, \ldots, b_5$. Attach your code to this assignment, 
write down the coefficients obtained, and plot the \emph{residuals}. You are encouraged to use
MATLAB or Python to solve this problem. Please attach all the problem
related work in a single place in your assignment (in other words,
arrange your writeup carefully so that we can find all the information
in one place).

\noindent\textbf{Hint:} First the problem needs to be posed as a
regression problem of the form $A \mathbf{x} \simeq \mathbf{b}$.
Identify what $A, \mathbf{b}$ and $\mathbf{x}$ are for this problem.


\bigskip

\noindent\textbf{P3 (25 points).} Consider the linear programming problem below:

\[ \begin{array}{llllllllll}
\max & -2 x_1 & - 3 x_2 & - x_3 & - x_4 &  & + x_6 \\[6pt]
\mathsf{s.t.} & x_1 & - x_2 &   &   &  & - x_6 & \leq 3\\[3pt]
&  x_1 &    &     &    -x_4 &   -x_5  &  & \leq -1 \\[3pt]
&  &  & - x_3 & & & - x_6 & \leq -2 \\[3pt]
&  & -x_2 & -x_3 & + x_4 & & + x_6 & \leq 4\\[3pt]
& x_1 & & + x_3 & & + x_5 & + x_6 & \leq 6\\[3pt]
& -x_1 & &  & -x_4 & + x_5 & - x_6 & \leq -2 \\[3pt]
& \multicolumn{6}{r}{x_1,\ \ldots,\ x_6} & \geq 0\\[3pt]
\end{array}\]

We add slack variables $w_1, w_2, \ldots, w_6$ for the $6$
constraints. For your convenience, the problem data is provided
separately for easy cut and paste into your python/matlab/other code.

\noindent\textbf{(A)} Compute and write out the 
dictionaries for the following
set of basic variables. If no dictionaries exist, say why.

\begin{enumerate}
\item $\{ x_1, x_2, x_3, w_1, w_2, w_3 \}$.
\item $\{ x_2, x_2, x_5, w_3, w_5, w_6 \}$.
\item $\{x_1, x_2, x_6, w_{4}, w_5, w_6 \}$.
\end{enumerate}


\noindent\textbf{(B)} Perform one step of the revised simplex method for the dictionary with the basis:
\[ \{x_3, x_4, x_5, w_1, w_2, w_6 \} \]

\begin{enumerate}
\item Compute the constant column and objective rows of this dictionary.
\item Choose the entering variable with the largest coefficient.
\item Set up the equations to construct the column for the entering
  variable.
\item Choose the leaving variable.
\item Write down the basic variables in the next dictionary.
\end{enumerate}

You may use MATLAB or Python to perform the matrix calculations
required. You are encouraged to code up this process as this will be
important for your programming assignment.


\noindent\textbf{(C)} Write down the final dictionary for the
problem. First you may want to solve the problem using your favorite
Simplex solver and work out what the basic variables should be from
your solution.


\noindent\textbf{(D)} We wish to now update the objective function to
a new function 
\[  (-2 x_1  - 3 x_2  - x_3  - x_4 + x_6 )  + \mu ( x_1 + x_2 +
  x_6) \]
where $\mu$ is a parameter. Write down the range of values of $\mu$
for which the final dictionary from part (D) continues to remain final
as the objective is changed.

\noindent\textbf{(E)} Going back to (C), we wish to update the right
hand sides of the original problem to 
\[ \left( \begin{array}{l} 3 \\ -1 \\ -2 \\ 4 \\ 6 \\ -2 \end{array}
  \right) + \lambda \left( \begin{array}{l}
1 \\ 1\\ 0 \\ 0\\ 0 \\ -1 \end{array} \right) \]

Calculate the range of values for $\lambda$ for which, the dictionary
in (C) remains feasible.







\end{document}
